---
title: "Estimation of potentially informative sites"
author: "Viacheslav Shalisko"
date: "1/4/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ape)
library(ips)
```

# Prepare data

* The inpud data should be alignment in NEXUS format
* The missing data should be N, not ?, if there are some ? they will be replaced by N
* The gaps are treated as significant characters, so the variable sites count may include gaps (except the trailing gaps that are considered as N)
* If you need to count ony variable sites in the sense of nucleotide substitutions, excluding gaps, replace all gaps by N



```{r}
my_alignment <- read.nex("10_regions_NEW2 - Alineamiento.nex.txt")
str(my_alignment)
print(my_alignment)

my_alignment_matrix <- as.character(my_alignment)

str(my_alignment_matrix)
my_alignment_matrix[my_alignment_matrix == '?'] <- 'n'
my_alignment_matrix[my_alignment_matrix == '-'] <- 'n'

my_alignment_ok <- as.DNAbin(my_alignment_matrix)

my_alignment_ok <- del.rowgapsonly(my_alignment_ok)
my_alignment_ok <- del.colgapsonly(my_alignment_ok)

str(my_alignment_ok)
print(my_alignment_ok)


my_accD <- my_alignment_ok[1:31,1:249]
my_ITS <- my_alignment_ok[1:31,250:1162]
my_matK <- my_alignment_ok[1:31,1163:2150]
my_psbAtrnH <- my_alignment_ok[1:31,2151:2635]
my_psbBTN <- my_alignment_ok[1:31,2636:3420]
my_rbcL <- my_alignment_ok[1:31,3421:4134]
my_rpl16 <- my_alignment_ok[1:31,4135:5073]
my_trnGS <- my_alignment_ok[1:31,5074:5605]
my_rpoB <- my_alignment_ok[1:31,5606:6228]
my_rpoC1 <- my_alignment_ok[1:31,6229:7072]

```

```{r fig.width=10, fig.height=10}
print("Without modification")
image(my_alignment, cex.lab = 0.5)
print("Gaps as N")
image(my_alignment_ok, cex.lab = 0.5)

image(my_accD, cex.lab = 0.5)
image(my_ITS, cex.lab = 0.5)
image(my_matK, cex.lab = 0.5)
image(my_psbAtrnH, cex.lab = 0.5)
image(my_psbBTN, cex.lab = 0.5)
image(my_rbcL, cex.lab = 0.5)
image(my_rpl16, cex.lab = 0.5)
image(my_trnGS, cex.lab = 0.5)
image(my_rpoB, cex.lab = 0.5)
image(my_rpoC1, cex.lab = 0.5)
```


# Whole alignment

```{r}
dim(my_alignment)[2]
length(seg.sites(my_alignment, strict = FALSE, trailingGapsAsN = TRUE))
pis(my_alignment, what = "absolute", use.ambiguities = FALSE)
pis(my_alignment, what = "fraction", use.ambiguities = FALSE)
pis(my_alignment, what = "index", use.ambiguities = FALSE)
base.freq(my_alignment)

dim(my_alignment_ok)[2]
length(seg.sites(my_alignment_ok, strict = FALSE, trailingGapsAsN = TRUE))
pis(my_alignment_ok, what = "absolute", use.ambiguities = FALSE)
pis(my_alignment_ok, what = "fraction", use.ambiguities = FALSE)
pis(my_alignment_ok, what = "index", use.ambiguities = FALSE)
base.freq(my_alignment_ok)
```

# matK

```{r}
# my_matK <- read.nex("Cedrela_matK.nex")
print(my_matK)
dim(my_matK)[2]
length(seg.sites(my_matK, strict = FALSE, trailingGapsAsN = TRUE))
pis(my_matK, what = "absolute", use.ambiguities = FALSE)
```

# rpl16

```{r}
# my_rpl16 <- read.nex("Cedrela_rpl16.nex")
print(my_rpl16)
dim(my_rpl16)[2]
length(seg.sites(my_rpl16, strict = FALSE, trailingGapsAsN = TRUE))
pis(my_rpl16, what = "absolute", use.ambiguities = FALSE)
```

# psbA-trnH

```{r}
# my_psbAtrnH <- read.nex("Cedrela_psbA-trnH.nex")
print(my_psbAtrnH)
dim(my_psbAtrnH)[2]
length(seg.sites(my_psbAtrnH, strict = FALSE, trailingGapsAsN = TRUE))
pis(my_psbAtrnH, what = "absolute", use.ambiguities = FALSE)
```

# rpoC1

```{r}
# my_rpoC1 <- read.nex("Cedrela_rpoC1.nex")
print(my_rpoC1)
dim(my_rpoC1)[2]
length(seg.sites(my_rpoC1, strict = FALSE, trailingGapsAsN = TRUE))
pis(my_rpoC1, what = "absolute", use.ambiguities = FALSE)
```

# psbB-T-N

```{r}
# my_psbBTN <- read.nex("Cedrela_psbB-T-N.nex")
print(my_psbBTN)
dim(my_psbBTN)[2]
length(seg.sites(my_psbBTN, strict = FALSE, trailingGapsAsN = TRUE))
pis(my_psbBTN, what = "absolute", use.ambiguities = FALSE)
```

# accD

```{r}
# my_accD <- read.nex("Cedrela_accD.nex")
print(my_accD)
dim(my_accD)[2]
length(seg.sites(my_accD, strict = FALSE, trailingGapsAsN = TRUE))
pis(my_accD, what = "absolute", use.ambiguities = FALSE)
```

# rpoB

```{r}
# my_rpoB <- read.nex("Cedrela_rpoB.nex")
print(my_rpoB)
dim(my_rpoB)[2]
length(seg.sites(my_rpoB, strict = FALSE, trailingGapsAsN = TRUE))
pis(my_rpoB, what = "absolute", use.ambiguities = FALSE)
```

# rbcLa

```{r}
# my_rbcL <- read.nex("Cedrela_rbcL.nex")
print(my_rbcL)
dim(my_rbcL)[2]
length(seg.sites(my_rbcL, strict = FALSE, trailingGapsAsN = TRUE))
pis(my_rbcL, what = "absolute", use.ambiguities = FALSE)
```

# trnS-G

```{r}
# my_trnGS <- read.nex("Cedrela_trnS-G.nex")
print(my_trnGS)
dim(my_trnGS)[2]
length(seg.sites(my_trnGS, strict = FALSE, trailingGapsAsN = TRUE))
pis(my_trnGS, what = "absolute", use.ambiguities = FALSE)
```

# ITS

```{r}
# my_ITS <- read.nex("Cedrela_ITS.nex")
print(my_ITS)
dim(my_ITS)[2]
length(seg.sites(my_ITS, strict = FALSE, trailingGapsAsN = TRUE))
pis(my_ITS, what = "absolute", use.ambiguities = FALSE)
```


